---
title: "General PRK interactome description"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: flatly
---

# Setup 

Libraries required
```{r libraries, message=FALSE, warning=FALSE}
library(dplyr)
library(here)
library(factoextra)
library(tibble)
library(tidyr)
library(UpSetR)
library(HGNChelper)
```

# Data input

On the input, there is an exported table from KNIME with transformed protein intensities and results of the statistical test.

```{r data-input, message=FALSE, warning=FALSE}
data <- read.delim(here('data', '20240502_report.txt'))
subset <- data %>%
  select(UniProt_Gene.Names..primary., UniProt_Protein.ID, ends_with("_imp"), ends_with("adj.P.val"), ends_with("logFC"))

colnames(subset) <- gsub("_log2_norm_imp", "", colnames(subset))
colnames(subset) <- gsub("PG.MaxLFQ.", "", colnames(subset))
colnames(subset) <- gsub("LIMMA_DIA_", "", colnames(subset))

subset <- subset %>%
  rename('GENEID' = 'UniProt_Gene.Names..primary.',
         'PROTID' = 'UniProt_Protein.ID',
         'CTRL_1' = '1_REP2_CTRL' ,
         'PRK1_1' = '2_REP2_PRK1' ,
         'PRK2_1' = '3_REP2_PRK2' ,
         'PRK3_1' = '4_REP2_PRK3' ,
         'CTRL_2' = '5_REP3_CTRL' ,
         'PRK1_2' = '6_REP3_PRK1' ,
         'PRK2_2' = '7_REP3_PRK2' ,
         'PRK3_2' = '8_REP3_PRK3' ,
         'CTRL_3' = '9_REP4_CTRL' ,
         'PRK1_3' = '10_REP4_PRK1' ,
         'PRK2_3' = '11_REP4_PRK2' ,
         'PRK3_3' = '12_REP4_PRK3' ,
         'CTRL_4' = '13_REP5_CTRL' ,
         'PRK1_4' = '14_REP5_PRK1' ,
         'PRK2_4' = '15_REP5_PRK2' ,
         'PRK3_4' = '16_REP5_PRK3' ,
         'CTRL_5' = '17_REP6_CTRL' ,
         'PRK1_5' = '18_REP6_PRK1' ,
         'PRK2_5' = '19_REP6_PRK2' ,
         'PRK3_5' = '20_REP6_PRK3')
```

# Perform a PCA on samples

```{r PCA, message=FALSE, warning=FALSE}
tmp <- subset %>%
  select(GENEID, CTRL_1:PRK3_5) %>%
  rowwise() %>%
  mutate(most_variable = sd(c_across(CTRL_1:PRK3_5))) %>%
  arrange(desc(most_variable)) %>%
  ungroup() %>%
  slice_head(n = 500) %>%
  select(-most_variable) %>%
  column_to_rownames(var = "GENEID") %>%
  t() %>%
  as.data.frame() %>%
  mutate(condition = c("CTRL", "PRK1", "PRK2", "PRK3",
                       "CTRL", "PRK1", "PRK2", "PRK3",
                       "CTRL", "PRK1", "PRK2", "PRK3",
                       "CTRL", "PRK1", "PRK2", "PRK3",
                       "CTRL", "PRK1", "PRK2", "PRK3")) 

 prcomp(tmp[, -501], center = TRUE, scale = FALSE) %>%
    fviz_pca_ind(habillage = tmp$condition, palette = c( "#F39B7FFF",  "#94C96E", "#315D00", "#90005D")) 
```

# Number of upregulated proteins per bait

Find which proteins are upregulated based on threshold of logFC > 1 and adjusted p-value < 0.05
```{r upregulated-proteins, message=FALSE, warning=FALSE}
subset <- subset %>%
  mutate(up_PRK1 = ifelse(PRK1.CTRL.adj.P.Val < 0.05 & PRK1.CTRL.logFC > 1, 1, 0)) %>%
  mutate(up_PRK2 = ifelse(PRK2.CTRL.adj.P.Val < 0.05 & PRK2.CTRL.logFC > 1, 1, 0)) %>%
  mutate(up_PRK3 = ifelse(PRK3.CTRL.adj.P.Val < 0.05 & PRK3.CTRL.logFC > 1, 1, 0))

```

Plot the barplot of upregulated proteins per bait
```{r upregulated-proteins-per-bait, message=FALSE, warning=FALSE}
subset %>%
  select(GENEID, starts_with("up")) %>%
  pivot_longer(cols = up_PRK1:up_PRK3, names_to = "bait", values_to = "upregulated") %>%
  mutate(bait = gsub("up_", "", bait)) %>%
  group_by(bait) %>%
  summarise(n_upregulated = sum(upregulated)) %>%
  ggplot(aes(x = reorder(bait, -n_upregulated), y = n_upregulated, fill = bait)) +
  geom_bar(stat = "identity", show.legend = FALSE) +
  theme_minimal(base_size = 17)+
  labs(x = "Bait", y = "Number of upregulated preys") +
  scale_fill_manual(values = c(  "#94C96E", "#315D00", "#90005D"))
```

# UpSet plot of upregulated proteins

UpSet plot of upregulated preys in particular baits.
```{r  upset-plot, message=FALSE, warning=FALSE}
subset %>%
  select(starts_with("up_")) %>%
  rename_with(~ sub("up_", "", .x), everything()) %>%
  upset(., order.by = "freq")
```

# Human Cell Map localization

In order to determine whether interactions of particular bait happen in a specific cellular compartment, upregulated preys were mapped to the human cell map experimental data.

```{r human-cell-map, message=FALSE, warning=FALSE}
# Data input for Human Cell Map:
hcm <- read.delim(here('data', 'humancellmap_20240502.txt'))

#genenames_newest <- getCurrentHumanMap()
#save(genenames_newest, file = here('data', "genenames_20240502.RData"))
load(here('data', "genenames_20240502.RData"))

# Update gene names in HCM
update_HCM <- checkGeneSymbols(hcm$symbol, species = "human", map = genenames_newest, unmapped.as.na = FALSE)
hcm <- left_join(hcm, update_HCM, by = c("symbol" = "x"))
hcm$Suggested.Symbol[hcm$Suggested.Symbol == "SEPTIN2 /// SEPTIN6"] <- "SEPTIN2"
hcm$Suggested.Symbol[hcm$Suggested.Symbol == "EPRS1 /// QARS1"] <- "QARS1"
hcm$Suggested.Symbol[hcm$Suggested.Symbol == "MPHOSPH6 /// PALS2"] <- "PALS2"
hcm$Suggested.Symbol[hcm$Suggested.Symbol == "MT-CO2 /// PTGS2"] <- "MT-CO2"

# Update gene names in TurboID
update_subset <- checkGeneSymbols(subset$GENEID, species = "human", map = genenames_newest, unmapped.as.na = FALSE)
subset <- left_join(subset, update_subset, by = c("GENEID" = "x"))
subset$Suggested.Symbol[subset$Suggested.Symbol == "MASP2 /// BRD8 /// C11orf58 /// KIFAP3"] <- "KIFAP3"
subset$Suggested.Symbol[subset$Suggested.Symbol == "C10orf88 /// SLC38A4"] <- "C10orf88"

save(subset, file=here("data", "data-preprocessed.Rdata"))
```

```{r human-cell-map2, message=FALSE, warning=FALSE}
# Map onto the Human Cell Map
data.hcm <- subset %>%
  left_join(., hcm %>% select(Suggested.Symbol, MMF.localization), by = c("Suggested.Symbol" = "Suggested.Symbol"))

data.hcm[is.na(data.hcm)] <- "-" #If protein was not mapped, use "-" instead

# Split the MMF.localization by "," in case there are multiple localizations mapped to the same protein
data.hcm <-  data.hcm %>%   
  mutate(NMF.loc = strsplit(as.character(MMF.localization), ", ")) %>% 
  unnest(NMF.loc)

x <-  subset %>%
  left_join(., hcm %>% select(Suggested.Symbol, MMF.localization), by = c("Suggested.Symbol" = "Suggested.Symbol"))

x[is.na(x)] <- "-" #If protein was not mapped, use "-" instead

# Split the MMF.localization by "," in case there are multiple localizations mapped to the same protein
x <-  x %>%   
  mutate(NMF.loc = strsplit(as.character(MMF.localization), ", ")) %>% 
  unnest(NMF.loc)

#write.csv(x, here('outputs', 'HumanCellMap_localization.csv'))

# Create long table
data.hcm <- data.hcm %>%
  select(Suggested.Symbol, starts_with("up_"), NMF.loc) %>%
  rename_with(~ sub("up_", "", .x), everything()) %>%
  pivot_longer(cols = PRK1:PRK3,
               names_to = "bait",
               values_to = "significant")

data.hcm %>%
  group_by(bait, NMF.loc) %>%
  summarise(n_preys = sum(significant)) %>%
  filter(n_preys != 0) %>%
  mutate(top_loc = case_when(
    n_preys == max(n_preys) ~ "max",
    TRUE ~ "")
  ) %>%
  mutate(top_localization = case_when(
    top_loc == "max" & bait == "PRK1" ~ "PRK1",
    top_loc == "max" & bait == "PRK2" ~ "PRK2",
    top_loc == "max" & bait == "PRK3" ~ "PRK3",
    TRUE ~ ""
      )) %>%
  ggplot(aes(x = NMF.loc, y = n_preys, fill = top_localization)) +
  geom_bar(stat = "identity")+
  theme_minimal()+
  theme(axis.text.x = element_text(angle = 45, hjust=1)) +
  labs(x = "Prey localization in Human Cell Map",
       y = "Number of preys") + 
  facet_grid(bait ~ .) + 
  scale_fill_manual(values = c("#b8B8B8",  "#94C96E", "#315D00", "#90005D" ))

# Filter out the "-" (non-mapped) proteins
data.hcm %>%
  group_by(bait, NMF.loc) %>%
  filter(NMF.loc != "-") %>%
  summarise(n_preys = sum(significant)) %>%
  filter(n_preys != 0) %>%
  mutate(top_loc = case_when(
    n_preys == max(n_preys) ~ "max",
    TRUE ~ "")
  ) %>%
  mutate(top_localization = case_when(
    top_loc == "max" & bait == "PRK1" ~ "PRK1",
    top_loc == "max" & bait == "PRK2" ~ "PRK2",
    top_loc == "max" & bait == "PRK3" ~ "PRK3",
    TRUE ~ ""
      )) %>%
  ggplot(aes(x = NMF.loc, y = n_preys, fill = top_localization)) +
  geom_bar(stat = "identity")+
  theme_minimal()+
  theme(axis.text.x = element_text(angle = 45, hjust=1)) +
  labs(x = "Prey localization in Human Cell Map",
       y = "Number of preys") + 
  facet_grid(bait ~ .) + 
  scale_fill_manual(values = c("#b8B8B8", "#94C96E", "#315D00", "#90005D"))

# Localization table
hcm.count <- data.hcm %>%
  group_by(bait, NMF.loc) %>%
  filter(NMF.loc != "-") %>%
  summarise(n_preys = sum(significant)) %>%
  filter(n_preys != 0) %>%
  mutate(top_loc = case_when(
    n_preys == max(n_preys) ~ "max",
    TRUE ~ ""))

#write.csv(hcm.count, here("outputs", "humancellmap_numbers.csv"))
```


